\documentclass[12pt,openany]{book}

\usepackage[a5paper,margin=0.65in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{enumitem}
\usepackage{tikz-cd}
\linespread{1.26}
\setlength{\emergencystretch}{2em}

\numberwithin{equation}{chapter}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[chapter]
\newtheorem{example}[definition]{Example}
\newtheorem{remark}[definition]{Remark}
\newtheorem{notation}[definition]{Notation}
\newtheorem{convention}[definition]{Convention}
\newtheorem{exercise}[definition]{Exercise}

\theoremstyle{plain}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{proposition}[definition]{Proposition}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}
\newtheorem{fact}[definition]{Commutative Algebra Fact}

\theoremstyle{remark}
\newtheorem*{claim}{Claim}

\newenvironment{solution}{\begin{proof}[Solution]}{\end{proof}}
\providecommand{\texorpdfstring}[2]{#1}

\renewcommand{\AA}{\mathbb{A}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\FF}{\mathcal{F}}
\newcommand{\GG}{\mathcal{G}}
\newcommand{\HH}{\mathcal{H}}
\newcommand{\II}{\mathcal{I}}
\newcommand{\JJ}{\mathcal{J}}
\newcommand{\LL}{\mathcal{L}}
\newcommand{\MM}{\mathcal{M}}
\newcommand{\NN}{\mathcal{N}}
\newcommand{\UU}{\mathfrak{U}}
\newcommand{\VV}{\mathfrak{V}}
\newcommand{\shHom}{\mathcal{H}om}
\newcommand{\dual}{\mathbin{\check{\ }}}
\newcommand{\tensor}{\otimes}
\newcommand{\xto}[1]{\xrightarrow{#1}}
\newcommand{\xfrom}[1]{\xleftarrow{#1}}
\newcommand{\cech}{\check{C}}
\newcommand{\cH}{\check{H}}
\newcommand{\kbar}{\overline{k}}

\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Proj}{Proj}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\Pic}{Pic}
\DeclareMathOperator{\Div}{Div}
\DeclareMathOperator{\CaDiv}{CaDiv}
\DeclareMathOperator{\Cl}{Cl}
\DeclareMathOperator{\Frac}{Frac}
\DeclareMathOperator{\Supp}{Supp}
\DeclareMathOperator{\Ass}{Ass}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\coker}{coker}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\codim}{codim}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\trdeg}{trdeg}
\DeclareMathOperator{\length}{length}
\DeclareMathOperator{\ord}{ord}
\DeclareMathOperator{\res}{res}
\DeclareMathOperator{\QCoh}{QCoh}
\DeclareMathOperator{\Coh}{Coh}
\DeclareMathOperator{\Der}{Der}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\Ann}{Ann}
\DeclareMathOperator{\Sym}{Sym}
\DeclareMathOperator{\Tor}{Tor}
\DeclareMathOperator{\Ext}{Ext}
\DeclareMathOperator{\depth}{depth}
\DeclareMathOperator{\Sing}{Sing}

\title{Algebraic Geometry Notes Toward Riemann--Roch}
\author{Renko}
\date{\today}

\begin{document}
\pagestyle{plain}
\maketitle
\tableofcontents

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}
These notes are organized toward one theorem: Riemann--Roch for smooth
projective curves.  The order of the chapters is deliberately linear.  A
result may use commutative algebra and material from earlier chapters, but not
from later chapters.  The exposition keeps the scheme-theoretic language
minimal: sheaves, schemes, morphisms, quasi-coherent sheaves, cohomology,
differentials, divisors, and finally curves.

Throughout, rings are commutative with identity.  A curve over a field means a
separated scheme of finite type over the field whose irreducible components
have dimension one.  In the final chapter, unless explicitly stated otherwise,
the curve is smooth, projective, and geometrically integral over the base
field.  This hypothesis implies \(H^0(X,\OO_X)=k\), so that the usual form of
Riemann--Roch has constant term \(1-g\).

The enlarged version follows the arrangement of Liu and Hartshorne where this
serves the present goal: affine and projective schemes, local properties,
finite and proper morphisms, coherent sheaves, Cech cohomology, differentials,
divisors, and the geometry of curves.  Exercises marked as used later are
included in the text with solutions.  Pure commutative algebra facts are stated
without proof; they are exactly the kind of algebraic input normally supplied
by a preceding commutative algebra course.

\input{sheaves}
\input{schemes}
\input{morphisms}
\input{modules}
\input{cohomology}
\input{differentials}
\input{divisors}
\input{curves}

\end{document}
